Survey of Cache Oblivious B-trees

### 2011-12-12 paper.tex
edits + latex fixes

\documentclass{style}

@@

@@

This paper presents a survey of cache-oblivious B-trees (CO B-trees). The CO
B-tree was first introduced by Bender et al. in 2000. It is an external memory
data structure - meaning that it optimizes the number of memory transfers it
performs rather than the number of CPU instructions that it executes. Unlike
other external memory data structures, like the widely used B-tree, the CO
B-tree does not require any information about the number of level in the
machine's memory hierarchy, the block size and capacity at each level, or the
relative speeds at each level. Hence, the CO B-tree is ``cache-oblivious''.
Despite this, the CO B-tree is able to achieve the same optimal search bound
as the B-tree and near optimal insertion and deletion bounds as well. In this
paper we first describe the original CO B-tree and explain how it is able to
achieve its memory performance. We then present and analyze three followup CO
B-tree data structures that improve upon weaknesses of the original: the
cache-oblivious string B-tree, which introduces support for variable length
keys, the cache-oblivious streaming B-tree, which incorporates improvements to
the insertion bound at the expense of the search bound, and lastly, the
cache-oblivious concurrent B-tree, which introduces support for concurrency.

@@

Most algorithms assume an idealized model of computation that consists of a
single CPU and memory as a flat array
that can be read and written. However, in reality memory is not a flat array
but rather a hierarchy from super fast on chip caches, to slower main memory,
to even slower disks. When a program requests data, the data is found in the
memory hierarchy and typically carried up the hierarchy to the fastest memory
layer possible, usually a cache. The CPU is then able to access it to perform
any necessary operations. Unfortunately, this cache often has a small capacity
and once it is full any new data requests cause some existing cached data to
be evicted. Transferring data from lower to higher levels in the memory
hierarchy often takes a long time. In fact, access times between the different
layers of the hierarchy can differ by many orders of magnitude and thus the
time to transfer data often dominates the overall runtime performance of an
algorithm. External memory algorithms are wary of this fact and aim to
optimizing the number of memory transfers it performs rather than the number
of instructions it executes. They often try to exploit the fact the memory
hierarchies transfer data in blocks rather than individual atomic units. Thus,
external memory algorithms are designed to not only limit the number of memory
transfers it performs but also preserve the locality of related data so that
when a memory access is performed, its cost is amortized.

@@

Since external memory algorithms are primarily concerned with accessing data,
it is no surprise that many external memory problems can be reduced to finding
a suitable data structure that efficiently implement certain operations. A
well known external memory data structure is the B-tree. At a high level, the
B-tree works by making each one of its nodes the same size as a memory block
in a particular level of the hierarchy. Furthermore, the fanout in the tree
(the number of children of a node) is proportional to the block size as well.
The B-tree incurs $O(log(N))$ memory transfers for \Search{}, \Insert{}, and
\Delete{} and is widely used in databases and filesystems.

@@

The B-tree has two theoretical shortcomings. First, the B-tree is analyzed in
a two-level memory model called the Disk Access Machine (DAM) \cite{Aggarwal}.
However, in reality, computing architectures typically have memory hierarchies
that are much larger than two. The argument for the DAM model is that in
practice only one level in the memory hierarchy tends to dominate the
performance and thus a two-level memory model is sufficient for all intents
and purposes. However, this does mean that, theoretically, B-trees can begin
to incur an increasing number of memory transfers if the dominating level in
the hierarchy changes.

Additionally, the B-tree expects to know the block size of its memory layer in
advance. B-Trees are implemented with parameterized code that needs to be
changed for every level in the memory hierarchy. Needless to say, this tuning
can be very difficult and error prone. Additionally, tuning this parameter may
even be impossible in certain heterogeneous computing environments.

The shortcomings of the DAM model gave rise to the cache-oblivious (CO) model
of computation. In the CO model an algorithm has no information about the
memory hierarchy. It does not know anything about the speed or the block size
of any level. Conceptually, the CO model allows algorithm designers to reason
about the simple two-level model like DAM but also prove results for any
arbitrary multi-level memory hierarchy. This is because algorithms that
perform well in a CO model should, in theory, perform well regardless of
changes in block size, cache capacity, and cache speed. Consequently, CO
algorithm should gracefully scale to multi-level hierarchies as well.

@@

In \cite{BenderDemainColton} Bender et al. introduce the cache-oblivious
B-tree (CO B-tree). The CO B-tree is similar to the normal B-tree except that
it does not require explicit parameters about the memory hierarch.

The CO B-tree is able to achieve the optimal search bound of $\Theta(log(N))$
memory transfers, the optimal scan bound of $\Theta(N/B)$ memory transfers and
near optimal insertion and deletion bound of $\Theta(log_B N +
\frac{log^2{N}}{B})$. The CO B-tree was a seminal data structure that gave
rise to many others.

This paper provides a survey of the different types of CO B-trees and
discusses the various trade offs that they make. We start by analyzing the
original CO B-tree and explain how it works. This includes a detailed
description of the packed memory array (PMA) and strongly weight-balance
binary search trees (SWSBT) that are integral to the CO B-tree's memory
performance. Following this discussion we provide an analysis of the CO String
B-tree,a CO B-tree with variable length keys, the CO Streaming B-tree, a CO
B-tree that improves the cost of insertion at the expense of search, and the
CO Concurrent B-tree, which allows multiple processes to access the data at
the same time.

@@

Bender et al. first introduced a cache oblivious B-tree (CO B-tree) in
\cite{BenderDemainColton}. The CO B-tree achieves its number of memory
transfers by structuring its elements as a strongly weight-balanced search
tree (SWBST). The SWBST is then laid out into a packed memory array (PMA)
using a van Emde Boas (vEB) layout. The SWBST and PMA play crucial roles in
the other cache-oblivious data structures that were developed after the CO
B-tree and thus are described at length in the next two sections.

@@

The ultimate goal of a CO B-tree is to reduce the number of memory transfers
while supporting \Search{}, \Insert{}, and \Delete{} operations. Thus, at a
fundamental level the CO B-tree still needs to maintain its data in some
structure. That structure is a balanced search tree. The search tree is laid
out according to a van Emde Boas (vEB) layout \cite{veb1}, \cite{veb2}. The
idea of a vEB layout is fairly straight forward. Let $h$ denote the height of
a tree. the vEB layout splits the tree approximately at height $h/2$. This
splits the original tree into a ``top'' subtree and many ``bottom'' subtrees.
Let $A$ denote the top subtree and $B_1... B_k$ denote the bottom subtrees.
The vEB layout is defined recursively using the order: $A, B1, ... B_k$. This
layout is also shown in Figure~\ref{fig:veb}.

A nice consequence of using a vEB layout is that it is known that it will use
$O(log_B N)$ memory transfers for searches \cite{veb1}, \cite{veb2}. Thus, the
CO B-tree has already achieved the desired number of memory transfers for
search queries. It still remains to be shown that the CO B-tree achieves the
desired number of memory transfers for dynamic inserts and deletes.

It is known that the CO B-tree is going to use a balance search tree, but what
kind? This is where the strongly weight-balanced search tree becomes useful.
The weight of a node is $u$ is equal to the total number of $u$'s descendants
plus 1. It can also be thought of as the sum of the weights of $u$'s children
plus 1. Formally, $u$'s weight is: $w(u) = 1 + \sum_{v \in children} w(v)$.

Most weight-balanced search trees require that the heights of a node's left
and right subtrees are simply off by a constant factor. The SWBST enforces the
stronger property that a node at height $h$ in the tree (height is the node's
distance from the leaves; leaves have a height of 1) has $\Theta c^h$
descendants where $c$ is some constant. This property becomes important later
on when laying the tree out in memory (see Section~\ref{sec:layout}).

A known data structure that achieves this property is the ``weight-balanced
B-tree'' \cite{swbst}. The CO B-tree leverages this data structure to maintain
its elements. Before SWBST can be used, however, it remains to be proven that
it maintains the the $c^h$ descendants property even after inserts and
deletes.

The definition of a SWBST given in \cite{swbst} provides us with this bound
for the weight of a node:

\begin{equation}
\frac{c^{h-1}}{2} \leq w(u) \leq 2c^{h-1}
\end{equation}

Let us now consider insertions. To insert an element into a WSBST we search
down the tree and find the natural place to insert the element - just like in
any tree structure. After inserting the element into some node $w$, we have to
check to see if any of $w$'s ancestors have become unbalanced by having a
weight greater than $2^{d-1}$. If such an ancestor exists, it follows that
every node along the path between that ancestor and $w$ will be un balanced as
well. Thus, we start with $w$ and move up the tree.

When we split a node, we naturally need to determine which children belong to
which of the two nodes. We cannot just divide the children evenly because it
may cause another violation. Thus, instead, we find the longest sequence of
elements such that the weight of that subsequence is $\lceil w(u)/2 \rceil$.
By doing this, we maintain the bound that $\frac{c^{h-1}}{2} \leq w(u) \leq
2c^{h-1}$. We cascade these changes up the tree until all of the ancestors are
balanced.

Deletions are supported in the exact same way, except, we check if a node has
a weight $w(u) < \frac{c^{h-1}}{2}$. In this case, we will merge two neighbors
together and proceed up the tree. There is one edge case that needs to be
handled. After merging, you may produce a new node that has a weight larger
than the bounds. If this happens we immediately split the node again. This can
be thought of as stealing nodes from a neighbor.

We have now shown that the SWBST achieves the desired property that the number
of descendants of a node is $c^h$. We can now show how to lay this tree out in
memory.

@@

Now that we have our elements structured in a SWBST, we are still left with
mapping this tree to memory and proving that it achieves our desired number of
memory transfers. The packed memory array (PMA) was developed to solve this
problem \cite{BenderDemainColton}. There are three main costs that the PMA is
concerned with.

First, the PMA needs to maintain open space in its memory for newly inserted
items. The PMA is able to accomplish this using $O(1 + (log^2 N) / B)$ memory
transfers. Doing so may seem like attempting to achieve contradictory goals:
maintaining the locality of elements while keeping space available for new
inserts. The PMA uses a technique described in \cite{packedmemoryarray}. At a
high level, the PMA maintains windows into the array. When a window becomes
too full or too empty, the elements are spread out within the context of a
larger window. Two parameters are required, the window size and the threshold
for trigger the spreading out. However, these parameters are not related to
the block size of the memory level and thus do not violate the CO model of
computation. Overall, the PMA is able to keep extra space in $O(1 + (log^2 N)
/ B)$ memory transfers.

Second, the PMA needs to handle the splitting and merging caused by the SWBST
discussed in the previous section. The PMA is able to accomplish this using
$O(1 + (log N) /B)$ memory transfers. The proof is based on the fact that each
node has $c^h$ descendants (recall, this is the key property that SWBST
provided). Thus, merging or splitting a node will cause $c^h$ nodes to move,
requiring $O(1 + c^h/B)$ memory transfers. It follows from the definition of
SWBST in \cite{swbst} that when a node is rebalanced, it will require $c^h$
insertions or deletions until it is rebalanced again. Thus, the amortized cost
of each insertion and/or deletion is $O(1/B)$. When you split a node, it may
go all the way up the tree and thus overall it will incur $O((log N) / B)$
memory transfers.

Third, the PMA needs to handle updating the parent and child pointers and may
become invalidated because of how the nodes moved around. The PMA is able to
update the parent and child pointers. If the pointers are all close together
in memory (e.g. within $B$ distance of each other) it is not big deal because
only a constant number of memory transfers are required. However, if they are
further apart the PMA introduces buffer nodes. The buffer nodes don't actually
do anything. All they do is serve to ``stretch'' the vEB layout so that nodes
do not move far away from one another. By using these buffers the PMA is able
to achieve a the pointer update cost using $O(1+ logB / \sqrt{B}) log^2 N)$
memory transfers.

Of these three terms, the dominating factor is maintaining room for newly
inserted elements - $O((log^2N)/B)$. We know that the SWBST achieves
insertions and deletions in $O(log_B N)$ if it is cache-aware. Thus, adding
PMA makes the SWBST cache-oblivious at the cost of incurring an additional
$O((log^2N)/B)$ memory transfers.

Therefore, overall, the CO B-Tree achieves insertions and deletions in
$O(log_B N + (log^2N)/B)$ memory transfers. However, it still achieves the
optimal $O(log_B N)$ memory transfers for searches, matching the performance
of ordinary, cache-aware, B-trees.

@@

While B-trees achieve an optimal bound for the number of memory transfers for
search, they do not perform as well for updates and insertions. The
buffered-repository tree (BRT) is a common alternative to the B-tree that
achieves an improved amortized $O(log(N)/B)$ memory transfers for inserts
compared to the $O(log_{B+1}N)$ memory access of B-trees. Of course, the
B-tree and BRT are analyzed with the DMA model and are not cache-oblivious.

This search vs insertion trade-off also exists in the cache-oblivious model.
The original BO B-tree described in Section~\ref{sec:original} is designed to
optimize for the search in this trade off. However, Bender et al. propose two
data structures that and achieve the opposite end: the shuttle tree and the
cache-oblivious lookahead array (COLA) \cite{BenderFaFi07}. These two data
structures are refereed to as ``streaming B-trees'', indicating how they are
optimized for insertions at the cost of search.

@@

The shuttle tree is similar in layout and structure as the original CO B-Tree.
It uses a strongly weight-balanced search tree described previously (see
Section~\ref{sec:structure}) that is laid out in a vEB (see
Section~\ref{sec:veb}) layout into a PMA (see Section~\ref{sec:layout}). A
notable difference between the shuttle tree and the CO B-tree is that the CO
B-tree splits the tree at height $h/2$ when creating the vEB layout while the
shuttle tree splits at height $F_k$, where $F_k$ is the $kth$ Fibonacci number
that is stickily less than the height of the tree. Thus if a tree's height
happens to be a Fibonacci number, $F_k$, then the shuttle tree will choose to
split the tree at height $F_{k-1}$. The top and bottom ``halves'' produced by
this split are organized just like before and shown in Figure \ref{fig:veb}.

@@

At a high level, the shuttle tree improves the efficiency of insertion by
adding buffers to each node in the tree. In fact, shuttle tree nodes have a
linked list of buffers rather than just a single buffer like the BRT. Each of
these buffers is another shuttle tree. The a shuttle tree node enforces two
constraints: the length of its linked list as well as the overall height (and
thus capacity) of each of the buffered shuttle trees contained within its
linked list.

The length of a node's linked list is related to the node's height in the
tree. A node has a buffer for every time it the leaf of a sub trees that is
split when creating the vEB layout. This idea is illustrated in Figure
\ref{fig:buffers}.

The shuttle tree's node's buffers have a maximum capacity that is enforced by
bounding its height. The size of the buffers increases going across the linked
list. Thus, a node's first buffer is smaller than its last buffer. The maximum
height of a buffer is given by $F_{H(k)} = F_{k-2 \lceil log_{\phi} k \rceil}
$ where $\phi$ is the golden ratio and $k$ is chosen such that $F_k$ =
$\xi(h)$ and $\xi(x)$ is defined recursively. If $h$ is a Fibonacci number
then $\xi(h) = h$ and otherwise $\xi(h) = \xi(h-f)$ where f is the largest
Fibonacci number less than $h$. The maximum height of an individual buffer is
related to the size of the recursive subtree for which the node is a leaf.
Since the recursive subtrees are split based on the largest Fibonacci number
less than $h$ (as described previously) it follows that the buffer sizes are
also based on Fibonacci numbers.

@@

The operations that are described by Bender et al. for the shuttle tree are
\Search{}, \Insert{}, and \Scan{}. An explicit discussion on \Delete{} is
missing; however, it is assumed that it is symmetric to \Insert{}.

\Search{} and \Scan{} work virtually the same way as the CO B-tree. The only
difference is that you have to search a child node's buffer before following
the pointer down the tree.

\Insert{} is also similar to the CO B-tree except you buffer elements as you
add them. When you insert an element you move down the tree from the root to
leaves and stop when you find a node that has a buffer (note that because of
the nature of Fibonacci numbers that it is possible that a node does not have
any buffers). You then attempt to insert the element into the smallest buffer
in the node's linked list. If this insertion causes the buffer to overflow
(e.g. the height of the buffer exceeds its maximum), you move all of the
elements from this buffer to the next largest buffer. When the largest buffer
overflows you take all of its elements and ``shuttle'' them down the tree.
Eventually you will hit a leaf node and inserting here may cause the balance
constraints on the SWBST to become invalidated. The shuttle tree rebalances
itself just like any other SWBST: the leaf node is split into two and this
split cascades up the tree until balance is restored.

The shuttle tree also needs to take care of the buffers belonging to a split
node. When a node $u$ is split into two nodes $u_1$ and $u_2$ the shuttle tree
immediately allocates enough space for all of the $u_2$'s buffers ($u_1$
simply reuses the space allocated for $u$'s buffers). We then scan over $u$'s
buffers and find the ones rooted in $u_1$ and $u_2$ and move them
respectively. This can be achieved by a single scan.

The core problem of maintaining room for inserts, bounding the costs of
merging and splitting, and bounding the costs of updating parent and child
pointers in the shuttle tree is achieved using the same PMA used with the CO
B-tree. This is decried in detail in Section~\ref{sec:layout}.

@@

\Search{} uses $O(log_B N)$ memory transfers where $N = \Theta(c^{F_k})$ and
$c$ is your fanout. Note that $c$ is not treated as a constant since this is a
cache-oblivious data structure. This bound is achieved by searching over the
top and bottom recursive subtrees at each level when traversing the tree. A
key trick is that when you visit a node you only search its largest buffer
(that last buffer in the linked list for that level). The smaller buffers will
be visited eventually as you recurse down the tree. Eventually you will get
down to a single subtree that can fit into $O(1)$ blocks and thus can be
searched with $O(1)$ memory accesses. Since the fanout is proportional to our
block size it will take us $O(log_B N)$ block accesses until we get down to a
subtree and thus, by induction, the worst case runtime is $O(log_B N)$.

@@

\Scan{} uses an amortized $O(L/B)$ memory transfers to scan in $L$ consecutive
elements. This cost obviously does not include the single \Search{} operation
that is required to find the first starting element of the range. Once that
first element's block is transferred, the shuttle tree proceeds by traversing
the tree and scanning in elements. Since the vEB layout enforces the spatial
locality of nearby nodes in the tree, the shuttle tree only has to read in a
new block once every $B$ memory accesses. Thus, each individual memory access
is is amortized to $1/B$ and the overall number of memory transfers is
$O(L/B)$.

>>>>

% preprint + pagenumbering for page numbers
\documentclass[preprint]{style}
\pagenumbering{arabic}

@@

% This gets rid of the block of whitespace on the first page
\makeatletter
\let\@copyrightspace\relax
\makeatother

@@

This paper presents a survey of cache-oblivious B-trees (CO B-trees). The CO
B-tree was first introduced by Bender et al \cite{BenderDemainColton}. It is
an external memory data structure; it optimizes the number of memory transfers
performed rather than the number of CPU instructions that executed. Unlike
other external memory data structures such as the widely used B-tree, the CO
B-tree does not have access to the parameter's of the machine's memory
hierarchy, including the block size/capacity at each level, or the relative
speeds. Hence, the CO B-tree is \textit{cache-oblivious}.  Despite this, the CO
B-tree is able to achieve the same provably optimal search bound as the B-tree
and near optimal insertion and deletion bounds. In this paper we first describe
the original CO B-tree and explain how it is able to achieve its performance.
We then present and analyze three followup CO B-tree data structures that
improve upon weaknesses of the original: the \textit{cache-oblivious string
B-tree}, which introduces support for variable length keys, the
\textit{cache-oblivious streaming B-tree}, which improves the insertion bound
at the expense of the search bound, and lastly, the \textit{cache-oblivious
concurrent B-tree}, which introduces support for concurrency.

@@

Most algorithms assume an idealized model of computation that consists of
memory as a flat byte-addressable array which can be read and modified in some
fixed time. However, in practice memory is not a flat array, but rather is
organized as a hierarchy consisting of fast on-chip caches (L1, L2, etc.), to
slower DRAM, to even slower disks. When a program requests data, the data is
located in the memory hierarchy and typically carried up the chain to the
fastest memory layer (closest to the CPU), usually a cache. Unfortunately, this
cache can only have a limited capacity, and once it is full any new data
requests necessitate some existing cached data to be evicted. Transferring data
from lower to higher levels in the memory hierarchy often takes an
exponentially long time; access times between the different layers can differ
by many orders of magnitude. Therefore, the time to transfer data often
dominates the overall runtime performance of an algorithm.

External memory algorithms are wary of this fact and seek to optimize the
number of memory transfers performed rather than the number of CPU instructions
executed. They often try to exploit the fact the memory hierarchies transfer
data in blocks rather than individual machine word units. Thus, external memory
algorithms are designed to not only limit the number of memory transfers, but
also preserve the locality of related data so that when a costly memory access
is performed, its cost is amortized.

@@

Since external memory algorithms are primarily concerned with accessing data,
it is no surprise that many external memory problems can be reduced to finding
a suitable data structure that efficiently implement certain operations. A well
known external memory data structure is the B-tree \cite{BayerM72}. The B-tree
works by making each one of its nodes the same size as a memory block in a
particular level of the hierarchy. Furthermore, the fanout in the tree (the
number of children of a node) is proportional to the block size as well.  The
B-tree incurs $O(\log_{B}{N})$\footnote{Throughout this paper, $\log$ with the
base omitted is base-$2$} memory transfers for \Search{}, \Insert{}, and
\Delete{} and is widely used in databases and file-systems.

@@

The B-tree has two theoretical shortcomings. First, the B-tree is analyzed in a
two-level memory model called the Disk Access Machine (DAM) \cite{Aggarwal}.
However, in practice, architectures typically have memory hierarchies that are
larger than two. The argument for the DAM model is that in practice, only one
level in the memory hierarchy tends to dominate the performance, and thus a
two-level memory model is sufficient.  However, this means that B-trees can
begin to incur an increasing number of memory transfers if the dominating level
in the hierarchy changes.

Additionally, the organization of a B-tree requires knowing the block size of
the memory layer in advance. B-trees are implemented with parameterized code
that needs to be changed for every level in the memory hierarchy. This tuning
can be very difficult and error prone. Additionally, tuning this parameter may
even be impossible in certain heterogeneous computing environments.

The shortcomings of the DAM model gave rise to the cache-oblivious (CO) model
of computation \cite{Frigo99}. In the CO model an algorithm has no information
about the memory hierarchy. It does not know anything about the speed or the
block size of any level. Conceptually, the CO model allows algorithm designers
to reason about the simple two-level model like DAM but also prove results for
any arbitrary multi-level memory hierarchy. This is because algorithms that
perform well in a CO model should, in theory, perform well regardless of
changes in block size, cache capacity, and cache speed. Consequently, CO
algorithm should gracefully scale to multi-level hierarchies as well.

@@

In \cite{BenderDemainColton}, Bender et al. introduce the cache-oblivious
B-tree (CO B-tree). The CO B-tree is similar to the normal B-tree except that
it does not require explicit parameters about the memory hierarchy.

The CO B-tree is able to achieve the optimal search bound of
$\Theta(\log_{B}{N})$ memory transfers, the optimal scan bound of $\Theta(N/B)$
memory transfers and near optimal insertion and deletion bound of 
$\Theta(\log_{B}{N} + \frac{\log^2{N}}{B})$. 
The CO B-tree was a seminal data structure that gave rise to many others.

This paper provides a survey of the different types of CO B-trees and discusses
the various trade offs that they make. We start by analyzing the original CO
B-tree and explain how it works. This includes a detailed description of the
packed memory array (PMA) and strongly weight-balance binary search trees
(SWBST) that are integral to the CO B-tree's memory performance. Following this
discussion we provide an analysis of the CO string B-tree (a CO B-tree with
variable length keys), the CO streaming B-tree (a CO B-tree that improves the
cost of insertion at the expense of search), and the CO concurrent B-tree (a CO
B-tree that allows multiple processes to efficiently access data concurrently).

@@

The CO B-tree optimizes the number of memory transfers by structuring its
elements as a strongly weight-balanced search tree (SWBST). The SWBST is then
laid out into a packed memory array (PMA) using a van Emde Boas (vEB) layout.
The SWBST and PMA play crucial roles in the other cache-oblivious data
structures that were developed after the CO B-tree and thus are described at
length in the next two sections.

@@

The ultimate goal of a CO B-tree is to reduce the number of memory transfers
while supporting \Search{}, \Insert{}, and \Delete{} operations. Thus, it is
not surprising that the CO B-tree maintains its data in a balanced search tree.
The search tree is laid out in memory according to a van Emde Boas (vEB) layout
\cite{veb1,veb2}. The idea of a vEB layout is fairly straight forward. Let $h$
denote the height of a tree. The vEB layout splits the tree approximately at
height $h/2$. This splits the original tree into a \textit{top} subtree and
many \textit{bottom} subtrees.  Let $A$ denote the top subtree and
$B_1,...,B_k$ denote the bottom subtrees.  The vEB layout is defined
recursively using the order: $A,B_1,...,B_k$. This layout is also shown in
Figure~\ref{fig:veb}.

A nice consequence of using a vEB layout is that it can be shown to use
$O(\log_{B}{N})$ memory transfers for searches \cite{veb1, veb2}. Thus, the CO
B-tree has already achieved the desired bound for search queries. It still
remains to be shown that the CO B-tree achieves the desired number of memory
transfers for dynamic inserts and deletes.

To support dynamic inserts/deletes, a \textit{strongly weight-balanced search
tree} (SWBST) is used \cite{swbst}.  The weight of a node $u$ is equal to the
total number of $u$'s descendants plus 1. It can also be thought of as the sum
of the weights of $u$'s children plus 1. Formally, $u$'s weight is $w(u) = 1 +
\sum_{v \in children} w(v)$.

Most weight-balanced search trees require that the heights of a node's left and
right subtrees differ by at most a constant factor. The SWBST enforces the
stronger property that a node at height $h$ in the tree (leaves have a height
of 1) has $\Theta(c^h)$ descendants where $c$ is some constant ($c > 4$). This
property becomes important later on when laying the tree out in memory (see
Section~\ref{sec:layout}).

The definition of a SWBST given in \cite{swbst} provides this bound for the
weight of a node $u$:
\begin{equation}
  \frac{c^{h-1}}{2} \leq w(u) \leq 2c^{h-1}
\end{equation}

Let us now consider insertions. To insert an element, search
down the tree and find the natural place to insert the element - just like in
any tree structure. After inserting the element into some node $w$, 
check to see if any of $w$'s ancestors have become unbalanced by having a
weight greater than $2c^{h-1}$. If such an ancestor exists, it follows that
every node along the path between that ancestor and $w$ will be unbalanced as
well. This requires splitting nodes to restore balance, starting from $w$
and moving up the tree.

When a node is split, we need to determine which children belong to which of
the two nodes. Dividing the children evenly is insufficient because it may
cause another violation. Instead, find the longest sequence of elements such
that the weight of the subsequence is $\lceil w(u)/2 \rceil$.  By doing this,
we maintain the bound that $\frac{c^{h-1}}{2} \leq w(u) \leq 2c^{h-1}$. Cascade
these changes up the tree until all of the ancestors are balanced.

Deletions are supported similarly, except we check if a node has a weight $w(u)
< \frac{c^{h-1}}{2}$. In this case, we merge two neighbors together and proceed
up the tree. There is one edge case that needs to be handled. The merging
procedure may produce a new node that has a weight larger than the bounds
allow. If this happens, immediately split the node again, recursively. This can
be thought of as stealing nodes from a neighbor.

We have now sketched that the SWBST has the desired property that the number of
descendants of a node is $\Theta(c^h)$. We now describe the tree's memory layout.

@@

Now that the elements structured in a SWBST, we are still left with the problem
of mapping this tree to memory and proving that it achieves the desired number
of memory transfers per operation. The packed memory array (PMA) was developed
to solve this problem \cite{BenderDemainColton}. There are three main costs
that the PMA is concerned with: \textit{inserts}, \textit{splits/merges}, and
\textit{pointer updates}.

First, the PMA needs to maintain open space in its memory for newly inserted
items. The PMA is able to handle an insertion into the SWBST using only $O(1 +
(\log^2{N}) / B)$ memory transfers (this is the cost of updating the memory
layout of the SWBST in the PMA, not the cost of updating the data structure
itself). Doing so may seem contradictory: maintaining the locality of elements
while keeping space available for new inserts. The PMA uses a technique
described in \cite{packedmemoryarray}. At a high level, the PMA maintains
windows into the array. When a window becomes too full or too empty, the
elements are spread out within the context of a larger window. Two parameters
are required: the window size and the threshold for triggering a spread out.
However, these parameters are not related to the block size of the memory level
and thus do not violate the CO model.

Second, the PMA needs to handle splits and merges caused by inserting into the
SWBST discussed in the previous section. The PMA is able to accomplish this
using $O(1 + (\log{N}) /B)$ memory transfers. The proof is based on the fact
that each node has $\Theta(c^h)$ descendants (recall, this is the key property
that SWBST provided). Thus, merging or splitting a node will cause
$\Theta(c^h)$ nodes to move, requiring $O(1 + c^h/B)$ memory transfers. It
follows from the definition of SWBST in \cite{swbst} that when a node is
rebalanced, it will require $\Theta(c^h)$ insertions or deletions until it is
rebalanced again. Thus, the amortized cost of each insertion and/or deletion is
$O(1/B)$. When a node is split, it may traverse all the way up the tree, and
thus the split/merge bound is $O((\log N) / B)$ memory transfers.

Third, the PMA needs to handle updating the parent and child pointers that may
become invalidated because of how the nodes moved around. If the pointers are
all packed together in memory (e.g. within $B$ distance of each other) then
only $O(1)$ memory transfers are required. However, if they are further apart
then a problem arises. To deal with this, buffer nodes are introduced.  Buffer
nodes ``stretch'' the vEB layout so that nodes do not move far away from one
another. By using buffers, the PMA is able to achieve a pointer update cost
using $O(1+ \frac{\log{B}}{\sqrt{B}}\log^2{N})$ memory transfers.

Of these three terms, the dominating factor is maintaining room for newly
inserted elements - $O((\log^2{N})/B)$. Therefore, overall the CO B-tree
achieves insertions and deletions in $O(\log_B{N} + (\log^2{N})/B)$ memory
transfers. It still achieves the optimal $O(\log_B{N})$ memory transfers for
searches, matching the performance of static cache-aware B-trees.

@@

While B-trees achieve an optimal bound for the number of memory transfers for
search, they do not perform as well for bulk updates and insertions. The
buffered-repository tree (BRT) is an alternative to the B-tree that achieves an
improved amortized $O(\log{\frac{N}{B}})$ memory transfers for inserts compared
to the $O(\log_{B}{N})$ memory access of B-trees. In practice, log-structured
merge (LSM) trees \cite{ONeil96} are used for high-insert workloads.

This search vs. insertion trade-off also exists in the cache-oblivious model.
The original CO B-tree described in Section~\ref{sec:original} is designed to
optimize for search in this trade off. However, Bender et al. propose two data
structures that and achieve the opposite end: the shuttle tree and the
cache-oblivious lookahead array (COLA) \cite{BenderFaFi07}. These two data
structures are refereed to as \textit{streaming B-trees}, indicating how they
are optimized for bulk insertions at the cost of search.

@@

The shuttle tree is similar in layout and structure as the original CO B-tree.
It uses a strongly weight-balanced search tree described previously (see
Section~\ref{sec:structure}) that is laid out in a vEB (see
Figure~\ref{fig:veb}) layout into a PMA (see Section~\ref{sec:layout}). A
notable difference between the shuttle tree and the CO B-tree is that the CO
B-tree splits the tree at height $h/2$ when creating the vEB layout, while the
shuttle tree splits at height $F_k$, where $F_k$ is the $k$-th Fibonacci number
that is strictly less than the height of the tree. Thus if a tree's height
happens to be a Fibonacci number $F_k$, then the shuttle tree will choose to
split the tree at height $F_{k-1}$. The top and bottom ``halves'' produced by
this split are organized just like before and shown in Figure \ref{fig:veb}.

@@

At a high level, the shuttle tree improves the efficiency of insertion by
adding buffers to each node in the tree. Shuttle tree nodes have a
linked list of buffers rather than just a single buffer like the BRT. Each of
these buffers is another shuttle tree. A shuttle tree node enforces two
constraints: the length of its linked list as well as the overall height (and
thus capacity) of each of the buffered shuttle trees contained within its
linked list are bounded, described below.

The length of a node's linked list is related to the node's height in the tree.
A node has a buffer for every time it is the leaf of a sub-tree that is split
when creating the vEB layout. This idea is illustrated in Figure
\ref{fig:buffers}.

The shuttle tree's node buffers have a maximum capacity that is enforced by
bounding its height. The size of the buffers increases going across the linked
list from left-to-right. Thus, a node's first buffer is smaller than its last
buffer. The maximum height of a buffer is given by $F_{H(k)} = F_{k-2 \lceil
\log_{\phi} k \rceil}$, where $\phi$ is the golden ratio and $k$ is chosen such
that $F_k$ = $\xi(h)$ and $\xi(x)$ is defined recursively: If $h$ is a
Fibonacci number then $\xi(h) = h$, otherwise $\xi(h) = \xi(h-f)$ where $f$ is
the largest Fibonacci number less than $h$. The maximum height of an individual
buffer is related to the size of the recursive subtree for which the node is a
leaf.  Since the recursive subtrees are split based on the largest Fibonacci
number less than $h$ (as described previously) it follows that the buffer sizes
are also a function of the Fibonacci numbers.

@@

The shuttle tree supports \Search{}, \Insert{}, and \Scan{}. An explicit
discussion on \Delete{} is missing; it is assumed that it is similar 
to \Insert{}.

\Search{} and \Scan{} work mostly the same way as the CO B-tree. The only
difference is that a node's buffer must be searched before following a
pointer down the tree.

\Insert{} is also similar to the CO B-tree except elements are buffered as they
are inserted. \Insert{} starts by move down the tree from the root to leaves
and stopping when a node that has a buffer is located (note that because of the
nature of Fibonacci numbers that it is possible that a node does not have any
buffers). The element is inserted into the smallest buffer in the node's linked
list. If this insertion causes the buffer to overflow (e.g. the height of the
buffer exceeds its maximum), all of the elements from the buffer are moved into
the next largest buffer (and so on, recursively).  When the largest buffer
overflows all of its elements are taken and \textit{shuttled} down the tree.
Eventually a leaf node will be reached and inserting may cause the balance
constraints on the SWBST to become invalidated. The shuttle tree rebalances
itself just like any other SWBST: the leaf node is split into two and this
split cascades up the tree until balance is restored.
%% TODO: explain what it means to be "shuttled" down the tree

The shuttle tree also needs to take care of the buffers when a node is split.
When a node $u$ is split into two nodes $u_1$ and $u_2$, the shuttle tree
immediately allocates enough space for all of the $u_2$'s buffers ($u_1$ simply
reuses the space allocated for $u$'s buffers). $u$'s buffers are then scanned
over (by a single scan) and the elements are placed in the new buffers
accordingly.

The problem of maintaining room for inserts, bounding the costs of
merging and splitting, and bounding the costs of updating parent and child
pointers in the shuttle tree is achieved using the same PMA used with the CO
B-tree. This is described in detail in Section~\ref{sec:layout}.

@@

\Search{} uses $O(\log_{B}{N})$ memory transfers where $N = \Theta(c^{F_k})$ and
$c$ is the fanout. This bound is achieved by searching over the
top and bottom recursive subtrees at each level when traversing the tree. A
key trick is that when you visit a node you only search its largest buffer
(the last buffer in the linked list for that level). The smaller buffers will
be visited eventually as you recurse down the tree. Eventually you will get
down to a single subtree that can fit into a single blocks and thus can be
searched with one memory accesses. Since the fanout is proportional to the
block size it will take us $O(\log_B N)$ block accesses until we get down to a
subtree and thus, by induction, the worst case runtime is $O(\log_B N)$.

% Why is fanout propto block-size? doesn't this violate cache-obliviousness

@@

\Scan{} uses an amortized $O(L/B)$ memory transfers to scan $L$ consecutive
elements. This cost does not include the single \Search{} operation that is
required to find the first starting element of the range. Once the first
element's block is transferred, the shuttle tree proceeds by traversing the
tree and scanning in elements in order. Since the vEB layout enforces the
spatial locality of nearby nodes in the tree, the shuttle tree only has to read
in a new block about once every $B$ elements. Thus, each individual memory
access is is amortized to $1/B$ and the overall number of memory transfers is
$O(L/B)$.
